---
title: "BDA-assignment PART I"
author: "Ermioni Athanasiadi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(MASS)
library(coda)
```

# Q4: Newton-Raphson-Algorithm
```{r}

newton_raphson <- function(n, 
                           y, 
                           t, 
                           alpha, 
                           gamma, 
                           delta, 
                           maxit, 
                           lambda_start, 
                           beta_start) {

  #####
  # this function is based on the gradient from Antoninas solution
  #####
  
  # starting values for parameters
  # NOTE: REPARAMETRIZED PARS ARE USED and denote lambda_tilde, beta_tilde
  lambda <- lambda_start
  beta <- beta_start
  # overall parameter vector theta (9x1 vector)
  theta <- c(lambda, beta)
  
  # initialize result matrix with 3 columns and nrow=maxit
  res <- matrix(0, maxit, 10)
  res[1, 1:8] <- lambda
  res[1, 9] <- beta
  
  for (i in 2:maxit) {
    
  
    # use estimates from previous step
    lambda <- res[i-1, 1:8]
    beta <- res[i-1, 9]
    
    # log posterior
    log_p <- sum((y + alpha)*lambda) - sum((t + exp(beta))*exp(lambda)) + (n*alpha + gamma)*beta - delta*exp(beta)
    res[1,10] <- log_p
    
    # gradient of log-posterior (vector of first derivatives)
    # lambda gradient vector of n=8
    grad_lambda <- (y +  alpha) - exp(lambda)*(t + exp(beta))
    # beta gradient of n=1
    grad_beta <- n*alpha + gamma - exp(beta)*sum(exp(lambda)) - delta*exp(beta)
  
    # build gradient vector
    grad <- c(grad_lambda, grad_beta)
    
    
    # 2nd order derivative of lambdas n=8
    g2_lambda <- -exp(lambda)*(t+exp(beta))
    g2_beta <- -exp(beta)*sum(exp(lambda))-delta*exp(beta)
    
    # cross derivatives
    cross_der <- -exp(lambda)*exp(beta)
    
    # build hessian of 9x9 dimensions
    hess <- matrix(0, 9, 9)
    diag(hess) <- c(g2_lambda, g2_beta)
    hess[1:8, 9] <- -exp(beta) * exp(lambda)
    hess[9, 1:8] <- -exp(beta) * exp(lambda)
    
      
    # N-R step
    theta <- c(lambda, beta)
    theta_new <- theta - solve(hess, grad) # res[i, 1:9] <- th[i-1,1:9]-1*solve(hess)%*%grad
    
    ## update parameters
    lambda <- theta_new[1:8]
    beta <- theta_new[9]
    
    # store new estimated pars
    res[i, ] <- c(lambda, beta, log_p)

    
    ## ensure that lambdas are > 0 ?
    # lambda[1] <- max(0.00001, lambda[1])
    # ...
    
    ## ensure that beta > 0 ?
    
  }
  
  res
  
}



```

```{r}
newton_raphson_old <- function(n, 
                           y, 
                           t, 
                           alpha, 
                           gamma, 
                           delta, 
                           maxit, 
                           lambda_start, 
                           beta_start) {

  #####
  # this function is based on the gradient from my old solution
  #####
  
  # starting values for parameters
  # NOTE: REPARAMETRIZED PARS ARE USED and denote lambda_tilde, beta_tilde
  lambda <- lambda_start
  beta <- beta_start
  # overall parameter vector theta (9x1 vector)
  theta <- c(lambda, beta)
  
  # initialize result matrix with 3 columns and nrow=maxit
  res <- matrix(0, maxit, 10)
  res[1, 1:8] <- lambda
  res[1, 9] <- beta
  
  for (i in 2:maxit) {
    
  
    # use estimates from previous step
    lambda <- res[i-1, 1:8]
    beta <- res[i-1, 9]
    
    # log posterior
    # first version: has algebraic mistakes
    # log_p <- sum(y*lambda*t - t*exp(lambda)) + beta*(n*alpha + gamma-1) + sum((alpha - 1)*lambda) -      exp(beta)*sum(exp(lambda)) - delta*exp(beta)
    # second version: 
    log_p <- sum(y*lambda - t*exp(lambda)) + sum((alpha - 1)*lambda) - exp(beta)*sum(exp(lambda)) + beta*(n*alpha + gamma) - delta*exp(beta)
    # third version: Antoninas solution
    # log_p <- sum((y + alpha)*lambda) - sum((t + exp(beta))*exp(lambda)) + (n*alpha + gamma)*beta - delta*exp(beta)
    res[1,10] <- log_p
    
    # gradient of log-posterior (vector of first derivatives)
    # lambda gradient vector of n=8
    # first version: grad_lambda <- (y*t +  alpha-1) - exp(lambda)*(t + exp(beta))
    grad_lambda <- (y +  alpha-1) - exp(lambda)*(t + exp(beta))
    # Antoninas solution: 
    # grad_lambda <- (y +  alpha) - exp(lambda)*(t + exp(beta))
    # beta gradient of n=1
    grad_beta <- n*alpha + gamma - 1 - exp(beta)*sum(exp(lambda)) - delta*exp(beta)
    # Antoninas solution: 
    # grad_beta <- n*alpha + gamma - exp(beta)*sum(exp(lambda)) - delta*exp(beta)
  
    # build gradient vector
    grad <- c(grad_lambda, grad_beta)
    
    
    # 2nd order derivative of lambdas n=8
    g2_lambda <- -exp(lambda)*(t+exp(beta))
    g2_beta <- -exp(beta)*sum(exp(lambda))-delta*exp(beta)
    
    # cross derivatives
    cross_der <- -exp(lambda)*exp(beta)
    
    # build hessian of 9x9 dimensions
    hess <- matrix(0, 9, 9)
    diag(hess) <- c(g2_lambda, g2_beta)
    hess[1:8, 9] <- -exp(beta) * exp(lambda)
    hess[9, 1:8] <- -exp(beta) * exp(lambda)
    
      
    # N-R step
    theta <- c(lambda, beta)
    theta_new <- theta - solve(hess, grad) # res[i, 1:9] <- th[i-1,1:9]-1*solve(hess)%*%grad
    
    ## update parameters
    lambda <- theta_new[1:8]
    beta <- theta_new[9]
    
    # store new estimated pars
    res[i, ] <- c(lambda, beta, log_p)

    
    ## ensure that exponentiated lambdas are > 0 ?
    # lambda[1] <- max(0.00001, lambda[1])
    # ...
  }
  
  res
  
}

```


```{r}

n <- 8
y <- c(4, 1, 5, 14, 3, 19, 7, 6)
t <- c(95, 16, 63, 126, 6, 32, 16, 19)

alpha <- 1.8
gamma <- 0.01
delta <- 1

maxit <- 80

lambda_start <- rep(0, 8)
beta_start <- 0


res <- newton_raphson(n, y, t, alpha, gamma, delta, maxit, 
                      lambda_start, beta_start)

```


```{r}
head(res)
summary(res)
```


The resulting $\theta$ vector is of the form
```{r}
round(colMeans(res[, 1:9]), 5)

round(res[80, 1:9], 5)

```


But needs to be backtransformed
```{r}
round(colMeans(exp(res[, 1:9])), 5)
```

Lets quickly compare with the results obtained based on the old solution (different gradient, same Hessian)
```{r}

res_old <- newton_raphson_old(n, y, t, alpha, gamma, delta, maxit, 
                      lambda_start, beta_start)

round(colMeans(exp(res_old[, 1:9])), 5)

```
Ok, we get fairly similar results. Lets stick with the newer version though.




# Q5: Laplace-Approximation
```{r}

theta <- res[80, 1:9]
lambda <- theta[1:8]
beta <- theta[9]

# build gradient
grad_lambda <- (y +  alpha) - exp(lambda)*(t + exp(beta))
grad_beta <- n*alpha + gamma - exp(beta)*sum(exp(lambda)) - delta*exp(beta)
grad <- c(grad_lambda, grad_beta)
  
# build hessian of 9x9 dimensions
hess <- matrix(0, 9, 9)
# 2nd order derivative of lambdas n=8
g2_lambda <- -exp(lambda)*(t+exp(beta))
g2_beta <- -exp(beta)*sum(exp(lambda))-delta*exp(beta)
# cross derivatives
cross_der <- -exp(lambda)*exp(beta)
diag(hess) <- c(g2_lambda, g2_beta)
hess[1:8, 9] <- -exp(beta) * exp(lambda)
hess[9, 1:8] <- -exp(beta) * exp(lambda)

# sample from the laplace approximation
set.seed(1995)
laplace_results <- MASS::mvrnorm(n = 10000, mu = theta, Sigma = solve(-hess))
hist(laplace_results, breaks = 50)
```

Ok, the laplace approximation looks fairly similiar to the distribution we obtain from MH (see below).


# Q6: Metropolis Hastings
```{r}
set.seed(1995)
#### random walk metropolis sampler ####


# observed data
y <- c(4, 1, 5, 14, 3, 19, 7, 6)
t <- c(95, 16, 63, 126, 6, 32, 16, 19)

alpha <- 1.8
gamma <- 0.01
delta <- 1

# MCMC settings
m <- 35000  # chain length
acc <- 0
burnin <- 5000

# function to estimate log posterior given the theta parameter
get_log_posterior <- function(theta) {

  lambda <- theta[1:8]   # log-lambdas
  beta <- theta[9]       # log-beta

  # old solution for log-posterior
  #sum((y + alpha - 1) * lambda) -
  #  sum(t * exp(lambda)) - exp(beta) * sum(exp(lambda)) + (length(y) * alpha + gamma - 1) * beta - delta * exp(beta)
  log_p <- sum((y + alpha)*lambda) - sum((t + exp(beta))*exp(lambda)) + (n*alpha + gamma)*beta - delta*exp(beta)
  log_p
}

# initialize results vector
mcmc_results <- matrix(NA, m, 10)

# starting values
theta_current <- rep(0, 9)
log_current <- get_log_posterior(theta_current)
mcmc_results[1, 1:9] <- theta_current
mcmc_results[1, 10] <- log_current


# tuning parameter for sd of proposal 
proposal_sd <- 0.283   # you may tune later

# MCMC loop
for (i in 2:m) {

  # random-walk proposal
  theta_cand <- theta_current + rnorm(9, 0, proposal_sd)

  log_cand <- get_log_posterior(theta_cand)
  logalpha <- log_cand - log_current

  # accept/reject
  if (log(runif(1)) < logalpha) {
    theta_current <- theta_cand
    log_current <- log_cand
    acc <- acc + 1
  }

  mcmc_results[i, 1:9] <- theta_current
  mcmc_results[i, 10] <- log_current
}



# acceptance rate --> should be 23%
acceptance <- acc/m
acceptance

# then remove burn-in period
mcmc_results <- mcmc_results[(burnin+1):m, ]


```
### Visualization
```{r}
# only for lambda1 -> samples for the posterior
par(mar = c(5, 5, 1, 2))
hist(mcmc_results[, 1], 
     xlab=expression(theta), ylab="Density", 
     main="",
     col="lightblue", border="grey", 
     breaks=70, 
     ylim=c(0, 1),  
     cex.axis=1.5, cex.lab=1.5, 
     cex.main=2, las=1, freq=FALSE)  
grid()  
```


```{r}
# line plot for theta samples
#png("plots/MH_line.png", width = 10, height = 8, units = "in", res = 300)
#par(mar = c(5, 5, 1, 2))
plot(mcmc_results[, 1:9], type="l", 
     xlab="Iteration", ylab=expression(theta), 
     ylim=c(-2,0.5),
     col="lightblue", lwd=2, 
     cex.axis=1.5, cex.lab=1.5, 
     cex.main=2, las=1)  
grid()  
#dev.off()


## histogram
#png("plots/MH_hist.png", width = 10, height = 8, units = "in", res = 300)
par(mar = c(5, 5, 1, 2))
hist(mcmc_results[, 1:9], 
     xlab=expression(theta), ylab="Density", 
     main="",
     col="lightblue", border="grey", 
     breaks=70, 
     ylim=c(0, 1),  
     cex.axis=1.5, cex.lab=1.5, 
     cex.main=2, las=1, freq=FALSE)  
grid()  
#dev.off()
```

Lets compare MH and laplace for lambda4:
```{r}

theta_post <- mcmc_results[, 1:9]

lambda4_laplace <- exp(laplace_results[,4])
lambda4_mcmc    <- exp(theta_post[,4])

par(mfrow=c(1,2))
hist(lambda4_mcmc, prob=TRUE, main="MCMC", col="grey")
hist(lambda4_laplace, prob=TRUE, main="Laplace", col="lightblue")

```

```{r}
set.seed(1995)

burnin <- 5000

# assume res is (M x 9): lambdas + beta
mcmc_samples <- as.data.frame(mcmc_results[, 1:9])
colnames(mcmc_samples) <- c(paste0("lambda", 1:8), "beta")

mcmc_samples$iter <- seq_len(nrow(mcmc_samples))

```

```{r}
mcmc_samples_long <- mcmc_samples %>%
  pivot_longer(cols = -iter,
               names_to = "parameter",
               values_to = "value")

ggplot(mcmc_samples_long, aes(x = iter, y = value, colour = parameter)) +
  geom_line(alpha = 0.8) +
  labs(x = "Iteration", y = expression(theta[i])) +
  theme_minimal(base_size = 14)

```
The values of the lambda vector are too close. Lets plot them separately in the next step.

```{r}
ggplot(mcmc_samples, aes(x = iter, y = beta)) +
  geom_line(colour = "lightblue", linewidth = 0.8) +
  labs(x = "Iteration", y = expression(beta)) +
  theme_minimal(base_size = 14)
```
```{r}
ggplot(mcmc_samples_long, aes(x = iter, y = value)) +
  geom_line(colour = "lightblue", linewidth = 0.6) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 3) +
  labs(x = "Iteration", y = "") +
  theme_minimal(base_size = 13)

```
#### Compare MH and Laplace
```{r}
laplace_samples <- as.data.frame(laplace_results)
colnames(laplace_samples) <- c(paste0("lambda", 1:8), "beta")
laplace_samples$iter <- seq_len(nrow(laplace_samples))

laplace_samples_long <- laplace_samples %>%
  pivot_longer(
    cols = -iter,
    names_to = "parameter",
    values_to = "Value"
  )

ggplot() +
  geom_histogram(
    data = mcmc_samples_long,
    aes(x = value, y = after_stat(density)),
    bins = 50,
    fill = "lightgrey",
    colour = "darkgrey",
    alpha = 0.5  # transparency mode
  ) +  
  geom_histogram(
    data = laplace_samples_long,
    aes(x = Value, y = after_stat(density)),
    bins = 50,
    fill = "lightgreen",
    colour = "darkgrey",
    alpha = 0.5  # transparency mode
  ) 
```

```{r}
ggplot() +
  geom_histogram(
    data = mcmc_samples_long,
    aes(x = value, y = after_stat(density)),
    bins = 50,
    fill = "lightgreen",
    colour = "lightgrey",
    alpha = 0.5  # transparency mode
  ) +  
    geom_density(
    data = laplace_samples_long,
    aes(x = Value),
    colour = "darkgrey"
    #alpha = 0.5  # transparency mode
    )
```

```{r}
laplace_dens <- laplace_samples_long %>%
  group_by(parameter) %>%
  do(
    {
      dens <- density(.$Value)
      # save the theta values and their density as dataframe
      df <- data.frame(x = dens$x, y = dens$y, iter = seq_len(nrow(as.data.frame(x = dens$x))))
      
    }
  )

laplace_dens <- laplace_dens %>%
  arrange(iter)
  
```

```{r}
ggplot() +
  geom_histogram(
    data = mcmc_samples_long,
    aes(x = value, y = after_stat(density)),
    bins = 50,
    fill = "lightgrey",
    colour = "darkgrey",
    alpha = 0.5  # transparency mode
  ) +
  geom_line(
    data = laplace_dens,
    mapping = aes(x = x, y = y, group = parameter),
    linewidth = 0.4,
    colour = "lightgreen"
  ) + 
  facet_wrap(~ parameter, scales = "free")



```


# Q7: Point estimate and credible interval for lambda4
```{r}

mcmc_pars_summ <- mcmc_samples_long %>%
  group_by(parameter) %>%
  summarise(mean = mean(value),
            median = median(value),
            sd = sd(value))
mcmc_pars_summ
```

```{r}

# first obtain posterior estimate of beta hyperparameter
beta_samples <- mcmc_samples_long %>% filter(parameter == "beta") %>% pull(value) %>% exp()


# posterior predictive mean of y4
y4_ppc <- rnbinom(
  n    = length(beta_samples),
  size = alpha + y[4],
  prob = (beta_samples + t[4]) / (beta_samples + 2*t[4])
)

mean(y4_ppc)

```


We want to obtain the posterior predictive mean of our infection rate for virus $v_4$:

$\mathbb{E}\left[Y_4|\lambda_4, D\right]$

Our joint posterior distribution is a Gamma distribution, i.e. P(y_i | lambda, beta) ~ Gamma(shape = alpha + SUM y_i, beta + n). However, the posterior predictive distribution is negative binomial.

Alternatively, we can obtain the estimate directly from our lambda samples:
```{r}
lambda4_samples <- mcmc_samples_long %>% filter(parameter == "lambda4") %>% pull(value) %>% exp()

exp_lambda4_samples <- t[4] * lambda4_samples

mean(exp_lambda4_samples)
round(mean(exp_lambda4_samples), 0)
```

Now lets compute the quantile-based credible interval:
```{r}
# equal tail interval
round(quantile(exp_lambda4_samples, probs = c(.025, .975)), 0)
```


# Q8: Predicted probability for p(lambda_4 > 0.11)
```{r}
lambda4_probs <- mcmc_samples_long %>% filter(parameter == "lambda4") %>% 
  mutate(pred_dummy = ifelse(exp(value) > 0.11, 1, 0))

lambda4_prob <- mcmc_samples_long %>% filter(parameter == "lambda4") %>%
    summarise(prob = mean(exp(value) > 0.11)) %>% pull(prob)  # use boolean check

```
