---
title: "BDA-assignment"
author: "Ermioni Athanasiadi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To clarify:
- Model 1: Base model only random intercepts or time effect too?
- Rescale Time for better interpretation of model estimators
- Lasso/Ridge regression: How close is "close to 0"?
- Lasso/Ridge: necessary to standardize all covariates (so that they are comparable?)
- what is the default threshold for lasso/ridge and do we need to change it?

To Do:
- finalize Q3 + 4 from (preliminary) models selected from Q2 
- write this into report
- start Code for Part II - Metropolis
 
- Check missing data pattern
- check how to obtain residual variance (ICC analogue for Bayesian GLMM)
- check loo::waic() function with coda.samples(model_run)
- check how we can transform the SEs that are in the logit model output from log-odds to probabilities

PART II
- Q1/Q2
- once u have summary of parameter, its "too late" to transform to other scale (e.g. log), u need to go back to samples and get log and then again summarize
-> e.g. for OR take ratio of all pairs and then summarize again from samples -> CHECK THIS

To Add to Report:
- HPD interval (Q1)



Issues to be resolved:
- clarify pow() and order of random effect parameter estimation
- add model with stronger priors
- convert par estimates to probabilities in model formula or after estimation?


Notes on Model 1:
- interpretation of the intercept in an only-intercept model (odds vs. odds ratio)


```{r}
library("R2OpenBUGS")
library("coda")
library("readr")
library(rjags)
library(R2jags)
library(tidyverse)
library(bayesplot)
library(loo)
library(GWalkR)
```

```{r}
getwd()
```


```{r}
df <- readRDS("data/GSPS.RData")

head(df)
glimpse(df)
summary(df)

df <- df %>%
  mutate(
    year_scaled = year-1984,
    id = as.numeric(as.factor(id)),  # because jags accepts only sequential integers
    year_fct = as.factor(year),
    working_fct = as.factor(working),
    female_fct = as.factor(female)
  )

X <- model.matrix(~ year, data = df)


```

# EDA
```{r}
gwalkr(df)
```


Which of the variables are baseline measures / time-constant?

- educ (years of schooling)
- 
```{r}

vars <- c("female", "handper")
var <- vars[2]
vars <- names(df)[-c(1,15:19)]  # without id and factored vars


for (i in seq_along(vars)) {
  var <- vars[i]
  time_varying_rows <- df %>% 
    select(value = .data[[var]], id) %>% 
    group_by(id) %>% 
    distinct(value) %>% 
    count() %>% 
    filter(n > 1) %>%
    rename(
      !!var := n
    )
  print(var)
  print(time_varying_rows)
}

```

Okey, the only real baseline variable that doesn't change over time in this dataset is the gender variable "female".

```{r}
df %>%
  summarise(
    
  )
```


# ***Question 1: Build & Fit Model***

We want to investigate employment status over time. Our first model takes the form

working ~ year + (1 | id)

only accounting for repeated measurements over time per individual.

Interestingly, our time variable 'year' contains subsequent years from 1984-1988, and then two "jumps", 1991, 1994. This makes treating the year variable as continuous a bit tricky.




#### Specify model in BUGS language for logistic mixed model
```{r}
jags_str <- "
model
{
  # Likelihood Level 1
  for (i in 1:Nobs) {
    logit(p[i]) <- beta[1] + beta[2] * year[i] + b0[id[i]]
    # start with simpler model?
    # logit(p[i] <- beta[1] + b0[id[i]])
    working[i] ~ dbern(p[i])

  }
 
  # Likelihood Level 2
  for(j in 1:Nsubj) {	
			b0[j] ~ dnorm(0,taub0)
  }
  
  # Priors
  sigmab0 ~ dunif(0,100)
	sigma2b0  <- pow(sigmab0,2)
	taub0 <- pow(sigma2b0,-1)

	
	for (k in 1:2){
		beta[k]  ~ dnorm(0.0,1.0E-2)  # weakly informative prior
	}

  # Parameters of interest
  # pi_year0 <- exp(alpha_0)/(1+exp(alpha_0))
  # pi_yearmean <- exp(alpha_0 + alpha_1)/(1+exp(alpha_0 + alpha_1))

}
"
```


#### ADD: Repeat with stronger Prior
```{r}

```


#### Prepare the infrastructure for model estimation 
```{r}
# Prepare the data
model_data <- list(
  Nobs     = nrow(df),
  Nsubj    = df %>% select(id) %>% unique() %>% nrow(),
  #x1      = X[,2]
  working  = df$working,
  year     = df$year,
  id       = df$id
)

# parameters to watch
model_parameters <- c(
  "beta", "sigmab0"
)


## can also be deleted --> if no convergence issues, then we can skip providing initial values
my.inits <- list(
  list(beta=c(0,0) , sigmab0=1),
  list(beta=c(0,0) , sigmab0=10),
  list(beta=c(0,0) , sigmab0=5)
)


```


#### Model estimation with rjags 
```{r}
# run the model
model_run <- jags(
  data = model_data,
  parameters.to.save = model_parameters,
  model.file = textConnection(jags_str),
  inits = my.inits,
  n.chains = 3,
  n.iter = 40000,
  n.burnin = 1000,
  n.thin = 2,
  jags.seed = 1010 # ADD SEED! for reproducibility
)


```
```{r eval=FALSE, include=FALSE}
model_run$BUGSoutput$summary

mcmc_samples <- as.mcmc(model_run)
bayesplot::mcmc_trace(mcmc_samples)
```

save model results for later
```{r}
save(model_run, file = "./results/model_run.RData")
```

read in the results:
```{r}
load("./results/model_run.RData")
```

## Redo analysis but store log likelihood
```{r}
jags_str <- "
model
{
  # Likelihood Level 1
  for (i in 1:Nobs) {
    logit(p[i]) <- beta[1] + beta[2] * year[i] + b0[id[i]]
    working[i] ~ dbern(p[i])
    
    # store the pointwise log-likelihood (for WAIC)
    loglik[i] <- logdensity.bern(working[i], p[i])
  }
 
  # Likelihood Level 2
  for(j in 1:Nsubj) {	
			b0[j] ~ dnorm(0,taub0)
  }
  
  # Priors
  # beta[1] ~ dnorm(0.0,0.01)
  # beta[2] ~ dnorm(0.0,0.01)
  
  sigmab0 ~ dunif(0,100)
	sigma2b0  <- pow(sigmab0,2)
	taub0 <- pow(sigma2b0,-1)

	
	for (k in 1:2){
		beta[k]  ~ dnorm(0.0,1.0E-2)
	}

  # Parameters of interest
  # pi_year0 <- exp(alpha_0)/(1+exp(alpha_0))
  # pi_yearmean <- exp(alpha_0 + alpha_1)/(1+exp(alpha_0 + alpha_1))

}
"
```

```{r}
# add loglik to parameters to watch
model_parameters <- c(
  "beta", "sigmab0", "loglik"
)
```

```{r}
# run the model
model_run2 <- jags(
  data = model_data,
  parameters.to.save = model_parameters,
  model.file = textConnection(jags_str),
  inits = my.inits,
  n.chains = 3,
  n.iter = 40000,
  n.burnin = 1000,
  n.thin = 2,
  jags.seed = 1010 # ADD SEED! for reproducibility
)

```

```{r}
mcmc_samples_all <- as.mcmc(model_run2)

mcmc_params <- !grepl("^loglik\\[", colnames(mcmc_samples_all[[1]]))  # [[1]] indicates the first of three chains

mcmc_samples_keep <- lapply(
  mcmc_samples_all,
  function(chain) chain[, mcmc_params]
)

class(mcmc_samples_keep) <- "mcmc.list"

```


Likelihood:
We use a Bernoulli-distribution for predicting the binary response variable (working status).

Priors:
- normal weakly informative priors for the fixed effects coefficients (intercept, time effect) with beta_i ~ N(0,0.1)
- uniform prior for the variance of the random intercept

#### Diagnostics and convergence checks
```{r}
summ <- model_run2$BUGSoutput$summary
summ[!grepl("^loglik\\[", rownames(summ)), ]

bayesplot::mcmc_trace(mcmc_samples_keep)
```


```{r}
summary(mcmc_samples_keep)
# plot(mcmc_samples)
gelman.diag(mcmc_samples_keep)
```


```{r}

mcmc.list <- as.mcmc.list(mcmc_samples_keep)
mcmc.as.data.frame <- function(mcmc){
    # input argument: the mcmc_samples object obtained from as.mcmc()
    df = as.data.frame(as.matrix(mcmc))
    df$chain <- rep(1:nchain(mcmc), each = nrow(mcmc[[1]]))

    return (df)
}

mcmc.df <- mcmc.as.data.frame(mcmc_samples_keep)
```

#### Obtain Inference summaries
-> Post-Process data with posterior draws and summary table
```{r}
beta_parameters <- c("beta[1]", "beta[2]")
beta_plot_df <- mcmc.df %>% 
    select(all_of(beta_parameters), "chain") %>%
    pivot_longer(
        cols = all_of(beta_parameters),
        names_to = "parameter",
        values_to = "value"
    )

beta_par_summary <- beta_plot_df %>%
    group_by(parameter) %>%
    summarize(
      mean = mean(value),
      median = median(value),
      # mode = ?,
      sd = sd(value)
    )

beta_hpddf <- HPDinterval(
  as.mcmc(
    do.call(
      rbind, mcmc_samples_keep[,beta_parameters]
    )
  )
) %>% as.data.frame()
beta_hpddf$parameter <- rownames(beta_hpddf)
rownames(beta_hpddf) <- NULL


beta_summary <- merge(
  beta_par_summary, beta_hpddf,
  by = 'parameter'
  )

rownames(beta_summary) <- beta_summary$parameter
rownames(beta_summary) <- c("$\\beta_0$", "$\\beta_1$")
beta_summary$parameter <- NULL
colnames(beta_summary) <- c('Mean', 'Median', 'SD', 'LL', 'UL')

beta_summary <- beta_summary %>% round(3)

beta_summary
```


Now obtain inference summary for random intercept:
```{r}
#########



#########
```

#### Add comparisons: Diff between Mean and Posterior Mean
```{r}

```


#### Interpretation
Large intercept, very small (negligible) time effect, no significant effect based in HPD interval. 

We cannot infer that time affects working status based on this model.


Random intercept Variance (Recheck):
~ 4, indicating that a lot of variation is captured 



# ***Question 2: Model 2-3***

*Ridge Priors*
= independent Gaussian priors with common variance

*Lasso Priors*
Laplace (double exponential) priors

Both can be applied for variable selection



Conventional Lasso/Ridge Regression:
Penalty is at the absolute vs. at squared value and can become 0
Bayesian Variant:
If you add a prior, the coefficient will never become exactly Zero, but very close to Zero




#### Idea: try out if coefficient estimates change much if stronger priors are applied
--> sort of Sensitivity analysis
```{r}

```




#### implement Ridge priors
```{r}

```

#### implement Lasso priors
```{r}

```



# ***Question 3: Model comparison***
#### compute WAIC and DIC

https://kevintshoemaker.github.io/NRES-746/LECTURE8.html 

*WAIC: Wattanabe-Akaike information criterion*

Is interpretable just like AIC, BIC and DIC and allows us to compare models fitted in a Bayesian framework via MCMC:

WAIC=−2(lppd−P_WAIC)
$$
WAIC = -2(lppd - P_{WAIC})
$$

$$
lppd = \sum_i log \int_\theta
P(y_i|\theta)P(\theta | y)d\theta
$$

$$
P_{WAIC} = \sum_i
Var[log(P(y_i|\theta_{post})]
$$

- integral is approximated by averaging across the posterior distribution of the likelihood of each data point --> need to compute likelihood of each data point in our MCMC model fitting algorithm
- $P_WAIC$ is a penalty term (= effective number of parameters). 
- Less complex models are more constrained (less flexible/variable in the likelihood of each data point across MCMC samples) --> less penalty, complex models --> greater penalty.

```{r}
# equaliza variable naming *************
#mcmc_samples <- as.mcmc(model_run)

loglik_names <- sprintf("loglik[%d]", 1:model_data$Nobs)

lls <- as.matrix(mcmc_samples_all)[, loglik_names]  # this stacks all three chains with row-binding for computing WAIC

waic_full <- waic(lls)
waic_full$estimates["waic",]


```

Let's do a small plausibility check, as the stacked matrix contains 58500 elements only.
```{r}
start(mcmc_samples_all[[1]])
end(mcmc_samples_all[[1]])
thin(mcmc_samples_all[[1]])
niter(mcmc_samples_all[[1]])

# same for other two chains

# rows_per_chain <- (end(mcmc_samples_all[[1]]) - start(mcmc_samples_all[[1]])) / thin(mcmc_samples_all[[1]]) + 1
```
Ok. Due to thinning and burn-in period, we end up with way less rows than the initially specified 40.000 iterations. 

*DIC: Deviance Information criterion*
#### compute DIC
```{r}
model_run$BUGSoutput$DIC

model_run2$BUGSoutput$DIC
```


```{r eval=FALSE, include=FALSE}

## doesnt work
dic.samples(model_run)

```

```{r}
# dics <- as.matrix(mcmc_samples_keep)[, "deviance", drop = FALSE]
mcmc_mat <- as.matrix(mcmc_samples_keep)
D <- mcmc_mat[, "deviance"]
mean(D)
```



While the WAIC gives us the summed per-observation deviance from the log-likelihoods, the DIC is the mean deviance across iterations.


This is another approach: resample from the whole model
obtained from: https://gist.github.com/oliviergimenez/68ad17910a62635ff6a062f8ec34292f 
```{r}
# THIS DOESNT RUN WITH RELOADED JAGS MODEL from .RData file -> must be recompiled
samples <- jags.samples(model_run$model, 
             c("WAIC", "deviance"), 
             type = "mean",
             n.iter = 40000,
             n.burnin = 1000,
             n.thin = 1
             )
```

```{r}

samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic.m0 <- round(c(waic = tmp[["waic"]], p_waic = tmp[["p_waic"]]), 1)
waic.m0



```
```{r}
str(model_run$BUGSoutput$sims.list)
head(model_run$BUGSoutput[[1]])

sims <- model_run$BUGSoutput$sims.list
D <- sims$deviance


```


# ***Question 4: Model evaluation***
#### generate predicted y_is from posterior draws
```{r}

colnames(mcmc.df)
# [1] "beta[1]"  "beta[2]"  "deviance" "sigmab0"  "chain" 

# take the model results and turn to dataframe
head(mcmc.df)
# as.mcmc
# mcmc.as.data.frame


# construct linear predictors for each draw with design matrices and parameter draws
betas <- grep("^beta", colnames(mcmc.df), value = T)
X_beta <- as.matrix(df$year) %*% t(as.matrix(mcmc.df[, betas]))
bs <- "sigmab0" 
b_i[j] <- rnorm(0, sd = sigmab0[j])
Z_b <- as.matrix(mcmc.df[, bs])
# eta[i, j] <- X_beta + 
eta[i, j] = beta0[j] + bs[j, i]

# compute the probability/ies with inverse link function
pred_prob[i, j] <- 1 / (1 + exp(-eta[i, j]))
nobs <- nrow(pred_prob)
nsamples <- ncol(pred_prob)

# sample predicted y with rbinom() with length of my dataframe (n=700) and probabilities as obtained from above
y_rep[i, j] <- rbinom(1, size = 1, prob = pred_prob[i, j])




```


```{r}

# mcmc.df = posterior draws (iterations x parameters)
# df = data frame with covariates (year) and n_obs = 700

nobs <- nrow(df)
nsamples <- nrow(mcmc.df)  # iterations * chains

# 1. Extract beta coefficients
betas <- grep("^beta", colnames(mcmc.df), value = TRUE)
X <- model.matrix(~ 1 + df$year)
beta <- t(as.matrix(mcmc.df[, betas]))
X_beta <- X %*% beta
dim(X)
#[1] 700   2
dim(beta)
#[1]     2 58500
dim(X_beta)  #  700 58500 = nobs x nsamples

# 2. Sample random intercepts per draw
sigmab0_draws <- mcmc.df$sigmab0
# For each posterior draw, sample a random intercept for each individual
set.seed(123)
Z_b <- matrix(rnorm(nobs * nsamples, mean = 0, sd = rep(sigmab0_draws, each = nobs)),
              nrow = nobs, ncol = nsamples)

# 3. Linear predictor
eta <- X_beta + Z_b

# 4. Inverse link to get predicted probabilities
pred_prob <- 1 / (1 + exp(-eta))

# 5. Generate posterior predictive outcomes
y_rep <- matrix(rbinom(nobs * nsamples, size = 1, prob = as.vector(pred_prob)),
                nrow = nobs, ncol = nsamples)

```

```{r}
# no extract column-wise summaries across draws
y_pred_mean <- rowMeans(y_rep)          # mean predicted probability per observation
y_pred_sd   <- apply(y_rep, 1, sd)      # uncertainty per observation

# Or for dataset-level summaries:
mean(y_rep)      # mean of all simulated outcomes
sd(y_rep)        # standard deviation
max(y_rep)
```
compare this to the observed values
```{r}
mean(df$working)
```



```{r}
y_obs <- mean(df$working)

# Summarize posterior predictive outcomes
y_rep_means <- colMeans(y_rep)  # mean of each simulated dataset

fig_post_pred_distr <- ggplot(data.frame(y_rep_means), aes(x = y_rep_means)) +
  geom_histogram(binwidth = 0.01, fill = "lightgrey", color = "grey") +
  geom_vline(xintercept = mean(y_obs), col = "red", lwd = 0.8) +
  labs(x = "Mean probability for employment", y = "Frequency"#,
       # title = "Posterior predictive distribution of predicted probability for working status"
       )


ggsave("./export/figures/q4_post_pred_distr.png", fig_post_pred_distr, heigh = 4, width = 4)
```

